Project Manager's Progress Report

### 2014-05-24 pnsac/newsletter/article_EditorsNotes.tex
Updated files for May 2014 newsletter.

Jim Riddoch came to the rescue with a story about a
successful\/disastrous flight test.  We like these kinds of
stories--to small and personal to turn up in the history books, but
they do pull you into real human experiences. There will be more of these.

%We also have another hair-raising flying story from Tim Timmins. A
%series of previous Timmins stories can be found through the same
%website, starting with the December 2010 issue.

>>>>

Jim Riddoch came to the rescue with a story about a
successful/disastrous flight test.  We like these kinds of
stories\textemdash too small and personal to turn up in the history books, but
they do pull you into real human experiences. There will be more of these.

%We also have another hair-raising flying story from Tim Timmins. A
%series of previous Timmins stories can be found through the same
%website, starting with the December 2010 issue
\vspace{12pt}

### 2014-05-24 pnsac/newsletter/article_Feature1.tex
Updated files for May 2014 newsletter.

\title{My Days at the Home of the Merlin Engine}
\subtitle{Part 4}
\author{Richard Lodge}

@@

One of the most mysterious departments at Rolls-Royce was the
Technical Publications Department. Word processing was unheard of in
the 1960s and every time an engine modification was made, it resulted
in much work documenting even a small change and distributing the
updated documentation to all engine users around the world. I never
spent any time working with the people in this department even though
there was a complete building devoted to the work. Compared with the
resources allocated to engine development, the technical publication
side of the business was very small and it always remained a mystery
to me.

Since retiring from my career as an accountant in Canada, I have
become active on the actual restoration work of the North
Star. Understandably, I have a special interest in working on the
Merlin engines. Nobody now working on the restoration has ever worked
on a Merlin when it was in service. It is therefore necessary at all
times to refer to the various publications and manuals we have managed
to accumulate for the North Star.

It is only now, actually working on the engine and referring
frequently to the exploded diagrams that I realize how complicated the
work of the Technical Publications Department must have been. The men
and women who worked there were largely unsung heroes in that it was
very tedious and unglamorous work which was of great importance for
the safe and trouble-free running of engines in service. I can
remember that I would hear from time to time of the difficulties of
ensuring that publication updates were actually delivered to many
remote places in the world and the considerable logistical
complications of ensuring that the packages arrived safely and
furthermore that the updates were actually inserted in the relevant
manuals, particularly as many of the engine users did not speak
English. I cannot remember now if any of the documentation was
translated into other languages although I have a recollection that we
employed people who could translate into French and Spanish.

After a couple of years I was given a new job in the Long Range
Planning Department. I was one of three accountants working in this
office under a very senior accountant. Here we looked at the costings
of the development work, the projected sales  and production costs on
new engine designs. This was the time when the Spey was going into
service and the the company had signed a fixed price contract for the
RB211 engines for the new Lockheed 1011. This was a major project,
championed by David Huddie the Chief Engineer. Computer spreadsheets
did not exist at that time - pencil and columnar paper were the order
of the day, together with efficient erasers. Long range forecasts of
25 years required many yearly columns and each time we produced
figures that the engineers did not like, we were told to rework the
numbers until we produced an acceptable answer. We were not limited in
the number of erasers we could use. The usual way of cutting costs was
to reduce the forecast development and testing time. Of course this
did not work in real life.

We were often dealing with the directors of R-R who lived on the top
floor of the Nightingale Road offices. Many of them were engineers and
they had little time for young bean counters. It was still a time of
optimism for the company. The huge contract for the Lockheed engines
had been signed and the company was still enjoying its reputation as
the company whose Merlin engines had saved Britain in the air
war. This was in 1965/6 and it was several years before our
projections of doom resulted in the bankruptcy of the
company. Fortunately I left the company in 1966 and was not involved
in what must have been a horrible process as a symbol of British
engineering excellence slowly went to the wall.

After the chaos caused by the demise of Rolls-Royce Limited, the
British government formed a new nationalized company to take over the
remains of the old one. This has been very much a success story and
the newly privatized Rolls-Royce is now a highly profitable and
innovative company. I look back on my days at R-R with nostalgia but I
now realize that much of the time we were all living in a fool's
paradise.

>>>>

\title{RAF Lightning with Over-Wing Tanks}
%\subtitle{Part 4}
\author{Jim Riddoch}

@@

During the early 60's Britain's Defence Department decided that the
new fleet of Lightning aircraft should have the capability to extend
its current sortie of 22 minutes to permit ferry to the Middle
East. In-flight refuelling had not worked successfully due to the
offset probe positioned under one wing. Accordingly BAC (British
Aircraft Corporation,Warton Division) was awarded a contract to
develop extended wing tanks that could be jettisoned if necessary.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{jim_riddoch.eps}
   %caption of the figure 
   \caption*{\small \em Jim Riddoch.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:jim_riddoch.eps}
\end{figure}

%% \begin{footnotesize}
%% [Image: {\normalfont\color{blue}\href{http://www.airteamimages.com/english-electric-lightning_XP693_united-kingdom---royal-air-force_43865.html}{English
%%     Electric Lightning with over-wing tanks}}]
%% \end{footnotesize}

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.5]{Eng3Intercooler.eps}
%   %caption of the figure 
%   \caption*{\small \em Garry installing the intercooler on engine 3.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:Eng3Intercooler}
%\end{figure}

Unfortunately the Lightning with its high speed wings had limited
capacity for wing tip tanks and already with armament pylons under
wing could not accommodate under-wing tanks, so the decision was made
to develop over-wing tanks secured with pylons through the wing
structure. These tanks would each carry 250 imperial gallons to
provide initial fuel transfer after take off and be jettisoned once
empty.

I was designated to be the Development Engineer to carry out the
development and testing of this design about 1963/64. The design
concept of these tanks had to include a balanced transfer system to
maintain aircraft C of G during take off and climb out, this meant 5
small interior sealed compartments pressurized with air to permit
sequential transfer. Also an emergency fuel jettison system prior to
tank wing ejection. As you can readily appreciate this was a very
challenging design and even more demanding for testing and
certification.

Once the initial test tanks were built they were dispatched to
Mechanical Test Department at Warton Lancashire, where I worked.
Needless to say the RAF were keen to witness the development and
testing of these over-wing tanks. A Wing Commander headed a team of
three Senior Techs to keep track of our progress. Fortunately the
Techs were very helpful and hands-on in working with us.  You can
imagine the challenges we faced checking the transfer sequences at
varying attitudes. Many a Saturday morning was spent testing and
verifying these change in attitudes and minimizing fuel transfer loss,
the end result being approximately 225 gallons of fuel transferred
from each wing tank depending on aircraft attitude and roll.

The next major design requirement was the emergency fuel jettison;
this was to be accomplished by air pressure from centre compartment to
rear of the tank where a jettisonable fairing was fastened by
explosive bolts and attached by a cord to a bung on the end of the
fuel ejection pipe. Talk about Heath Robinson device, this takes the
biscuit! Suffice to report that on the ground test of this
ill-conceived system with Chief Designer and my Chief Engineer and
many observers with cameras, the rear fairing departed into oblivion
and the cord broke, due to a miserable sub standard cord substituted
instead of a parachute cord as specified. The bung remained firmly
attached to the jettison pipe. It was about this time I first thought
of emigrating to Canada, in fact anywhere outside Britain!

Needless to say the cord was replaced, the ejection system proved
satisfactory and the tanks were certified for flight test. This was to
be done with one tank attached to the top of one wing and painted with
black and white checkered squares for easy recognition when ejected. I
always remember the Test Pilot, Peter Knight inquiring about the
likely occurrence of the tank separating from the wing and hitting the
tail of the aircraft. I assured him that ground testing of a model in
the wind tunnel had worked, in fact the small tank had disappeared
through the end of the tunnel and took some time to recover in the
neighbouring field.

To replicate the fuel system jettison some bright spark proposed
filling the tank with coloured water for better visibility. The
aircraft took off, circled Warton Aerodrome and made a low pass to
jettison the coloured water over the attentive crowd including chiefs
and dignitaries. Fortunately, the jettison of the tail fairing and
bung release worked perfectly and the water emerged in a colourful
shower.

The weather gods not being with us, the wind blew all the water over
the spectators covering their expensive clothing with a colourful dye!
Finally, the tank was ejected safely over Morecombe Bay and eventually
recovered and returned to BAC Warton amidst embarrassed and colourful
faces. As one might expect, this completely successful test left the
Wing Commander disgusted with the entire experiment. He cancelled the
project.

Any thoughts I had of joining the RAF completely disintegrated. Once
again, immigration to Canada looked much more inviting.

### 2014-05-24 pnsac/newsletter/article_Feature2.tex
Updated files for May 2014 newsletter.

\title{The Whispering Giant}
\author{Dave Lambeth}

@@

Twinkle, twinkle, old North Star,\\
With wrinkled skin and creaky
spar,\\
For years you’ve always been on
sked,\\
As many of our force you’ve sped\\
to distant lands beyond the skies,\\
you’ve never failed to see the rise\\
of distant suns or moonrise glare.\\
You’ve carried us with tender care\\
-back you came to take us home,\\
from old Japan or France or Rome\\
from England, Goose, and Kef,\\
We cannot hear you.\\
(‘cause we’re deaf)\\

>>>>

\title{Elise D\'{e}roche, Aviatrice}
\author{Bruce Grant, Fl\^{a}neur}

@@

Wandering about in Paris recently I came upon a plaque, bolted to a
wall at 61 Rue de la Verrerie in the 4th. Wandering about is my
favourite activity in Paris. Let others wait in the long lines at the
tourist sites (I've done all that) or plunder the shops in Faubourg
Saint Honor\'{e} (I'll never do that); I wander about.

So, who was this Elise D\'{e}roche a.k.a. Baronne Elisa Raymonde de
Laroche, Aviatrice?  I took the picture and resolved to look her up at
home.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{aviatrice.eps}
   %caption of the figure 
   %\caption*{\small \em Garry installing the intercooler on engine 3.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:aviatrice}
\end{figure}

Here's a brief outline of the story I encourage readers to look
for her online.  Many sites relate her story though with some
contradictions among the various accounts.

Born into the working class family of a plumber, Elise D\'{e}roche
became an artist and actress, took the stage name of Baronne Elisa
Raymonde de Laroche. She also became the "amie intime" of the artist
and aviator L\'{e}on Delagrange who introduced her to his friend Charles
Voisin, a builder of airplanes.

There is an interesting chronology that indicates a life full of
tragedy, a life that she lived with intensity and audacity: L\'{e}on
Delagrange died, January 1910 when the wing came off his Bl\'{e}riot
airplane. Elise recovered quickly from her grief and appealed to her
new ami intime Charles Voisin to teach her to fly. She won her pilot's
licence in March 1910. A crash in July 1910 put her in hospital with
18 fractures. She recovered and returned to flying.

In a car crash in September 1912, Charles Voisin was killed but
Elise survived. She maried the pilot Jacques Vial in February of
1915, her first marriage. During the war, she was not allowed to
fly but served in the military as a driver. In 1918 her only
child Andr\'{e} (with L\'{e}on Delagrange) died aged 15 of the Spanish
Flu. July 1919, she was killed in a plane crash while training to
become a test pilot. Audace? Mais oui!

La Baronne Elisa Raymonde de Laroche is buried in Cimeti\`{e}re
P\`{e}re Lachaise in Paris.

\vspace{10mm}

\begin{centering}
My candle burns at both ends;\\
It will not last the night;\\
But ah, my foes, and oh, my friends --\\
It gives a lovely light!\\
\vspace{2mm}
\begin{footnotesize}
Edna St. Vincent Millay
\end{footnotesize}

\end{centering}

### 2014-05-24 pnsac/newsletter/article_Feature3.tex
Updated files for May 2014 newsletter.

\title{Troop Seats and Extras}
\author{Nelson Smith}

@@

\section{Troop Seats}
\label{troopseats}

Seat back assemblies are completed for the troop seats. Thirty eight
units were produced. The assembling of the backs to the platforms to
produce triple and quad seats will be the next step. This will be done
over the winter.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{troopseats.eps}
   %caption of the figure 
   \caption*{\small \em New troop seats.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:troopseats}
\end{figure}

The prototype is progressing well. As previously stated, measurements
are a rarity with these projects and sometimes it takes longer to
figure out a plan than to do the job. A guesstimate for completion:
early December.

\section{Other Projects}
\label{otherporjects}

Proctective covers for the probes which protrude under the belly of
the North Star were made and installed. The foam pads previously
installed had done their time.

The First Officer's protective seat covers have been manufactured and
installed. This addition was required to shield the seat covers from
UV rays, dust and stains. Similar covers will be made for the
Captain's chair during the winter months.

>>>>

\title{The Whispering Giant}
\author{Dave Lambeth}

@@

\begin{centering}
Twinkle, twinkle, old North Star,\\
With wrinkled skin and creaky
spar,\\
For years you've always been on
sked,\\
As many of our force you've sped\\
to distant lands beyond the skies,\\
you've never failed to see the rise\\
of distant suns or moonrise glare.\\
You've carried us with tender care\\
-- back you came to take us home,\\
from old Japan or France or Rome\\
from England, Goose, and Kef.\\
We cannot hear you.\\
('cause we're deaf)\\
\end{centering}

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{jim_jung_2.eps}
   %caption of the figure 
   \caption*{\small \em A Whispering Giant (North Star 17515) at
     Resolute Bay NWT in about 1963, taken by Radio Operator Jim Jung.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:jim_jung}
\end{figure}

### 2014-05-24 pnsac/newsletter/article_PresReport.tex
Updated files for May 2014 newsletter.

Canada Day is now behind us.  We had an excellent day with a good
number of our volunteers showing off the North Star to the general
public. Judging by the activity around the plane all the day, we have
now developed a display format which is interesting to visitors.
Canada Day was the only day this year when the plane was on display to
the general public.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{canadaday_visitor.eps}
   %caption of the figure 
   \caption*{\small \em Showing off the North Star on Canada Day -- an
   impressed visitor examines the cockpit.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:canadaday}
\end{figure}

On June 27, in co-operation with the Museum, who hosted the event, we
showed off the plane to a group of donors attending the Donor
Appreciation Evening.  This was a successful evening which is likely
to be repeated.

In my last Notes from the President, I referred to plans for
celebrating the 10th anniversary of the founding of the Project North
Star Association.  This should have taken place around June 24.
Unfortunately, the event passed without us being able mark the
day. This was mainly due to situations beyond our control.

Our 2013 AGM was held on June 8, and all the existing directors were
re-elected.  This will enable the directors, who have now become
experienced, to concentrate on long-term matters.

The actual work on the restoration of the plane is going very well, as
reported elsewhere in this newsletter. The major issue facing the
Board in the next year will be to consider how to secure the future of
PNSAC, both in terms of volunteers and the funding of restoration
work.

We have now reached a point where it is necessary to update our
Memorandum of Understanding (MOU) with the Museum and our work plan
for completing the restoration of the plane.  Our volunteers have
already worked over 60,000 hours on the project and we expect to work
a similar number of hours before the restoration is completed.

There have been several informal meetings to discuss our future
operations and the opportunities available for new volunteers.  Within
the next few weeks we are expecting to meet with the Director General
of the Aviation Museum and members of his staff to work on the MOU
update.

Later this month we are meeting with senior management of the Museum
Foundation, the body charged with corporate fundraising for the Canada
Science and Technology Museums Corporation, to discuss how PNSAC can
work with the Foundation to raise funds for the North Star restoration
and the museums in general.

With the initiatives above, we are hoping in the next few months to
consolidate our position as an important organization working with the
Aviation Museum.

>>>>

Since my last Notes from the President in December 2013, restoration
work on the North Star has mostly been suspended. The Museum has
needed extra working space so that exhibits could be made ready for
the commemoration this year of the centenary of the start of the First
World War and the 70th anniversary on June 6 of D-Day in the Second
World War. These preparations have resulted in our normal working
areas being unavailable for North Star restoration work.

\begin{figure}[htbp]
  \vspace{2em}
  \centering
  %name of the graphic, without the path AND in EPS format:
  \includegraphics[scale=0.5]{richard_mtg.eps}
  %caption of the figure 
  \caption*{\small \em Richard at the Quarterly Meeting in April.}
  %label of the figure, which has to correspond to \ref{}:
  \label{fig:richard_mtg}
\end{figure}

During the shut down period some limited work has been continuing. Two
volunteers have been working in the cockpit of the plane and two other
volunteers have been doing some work on \#3 engine reassembly.

While the restoration work has been on hold, we have arranged three
monthly Volunteer Lunches at the Museum to enable the people who
normally work on the plane to meet and socialize. Although we would
all like to be working on the plane, we understand why the Museum has
had to ask us to temporarily suspend our activities. The Volunteer
Lunches are a way of keeping the team together and up-to-date on what
is happening.

At the last meeting of the PNSAC Board of Directors we agreed to
reschedule the Annual General Meeting from June each year to
September. It has always been difficult to find a good weekend in June
to hold the AGM and it is felt that a middle of September date would
be much better. Our corporate secretary, Roger Button, is aware of
this proposed change and has confirmed that the existing Board of
Directors can extend its term until the new AGM date.

Although there has been little restoration work done since December,
we are working with the Museum to get the project back on track again
in early June. We are also working to enable us to recruit new
volunteers to replace the several people who for various reasons have
had to give up active work.

One of the recent discussion areas has been in connection with
long-term funding of the restoration project. Everybody is aware that
funding from the Federal Government has been substantially reduced. We
have accepted that major  expenditures will have to be met from funds
raised by donations from friends of the Museum and members of our
Association. This is a significant challenge since everybody receives
many appeals throughout the year for donations from many different
deserving organizations.

During the next few months we hope to develop a revised restoration
work plan which will enable us to estimate the funds needed to
complete the project. We will then be able to target appeals for funds
to carry out specific parts of the restoration in the same way as we
did for our successful appeal for funds to make troop seats for the
interior of the plane.

### 2014-05-24 pnsac/newsletter/article_ProgReport.tex
Updated files for May 2014 newsletter.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{Eng3Intercooler.eps}
   %caption of the figure 
   \caption*{\small \em Garry installing the intercooler on engine 3.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:Eng3Intercooler}
\end{figure}

@@

While restoration work in the shop could not be continued, we
used the opportunity to install the crew seats and other restored
items to the cockpit.  The pilot seat was installed (first
officer seat was installed last year) then the railings and
safety board were installed behind the seats.  Next we installed
the radio officer and navigator tables, and the seats for these
positions.  The radio officer seat proved to be a particular
challenge, as the floor in this area is no longer flat, and some
modifications to the seat rails were necessary to allow the seat
to slide.  Finally the flight engineer’s jump seat was attached
to the railing.  With five crew seats installed, the cockpit is
very cramped.

@@

@@

The floor under this section was repaired by removing a section
of badly corroded “I” beam and fabricating a new one from several
pieces of aluminum which were riveted in place to support the
floor.  Once this repair was completed, new cushion flooring was
cut and glued to the crew lounge and the washroom floor, then
various pieces of floor trim were installed.  To install the crew
seat, it was necessary to remove the lower support section, since
the fully assembled seat could not easily fit through the narrow
doorway.  Once the seat was installed, we added the crew table
and bunk bed, along with the secure storage box.  We also
installed the curtain rod and new curtains to close off the bunk
bed. The installation was finished off by re-installing the door
between the lounge and the main cabin, so that this fully
restored area can be protected from dust and debris while the
main cabin undergoes restoration over the next two years.

>>>>

@@

While restoration work in the shop could not be continued, we
used the opportunity to install the crew seats and other restored
items to the cockpit.  The pilot seat was installed (first
officer seat was installed last year) then the railings and
safety board were installed behind the seats.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
   %caption of the figure 
   \caption*{\small \em Forward cockpit -- complete.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:fwd_cockpit_complete}
\end{figure}

Next we installed
the radio officer and navigator tables, and the seats for these
positions.  The radio officer seat proved to be a particular
challenge, as the floor in this area is no longer flat, and some
modifications to the seat rails were necessary to allow the seat
to slide.  Finally the flight engineer's jump seat was attached
to the railing.  With five crew seats installed, the cockpit is
very cramped.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{radio_position.eps}
   %caption of the figure 
   \caption*{\small \em Radio Officer's position showing the morse
     code key.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:radio_position}
\end{figure}

@@

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{cockpit_curtain.eps}
   %caption of the figure 
   \caption*{\small \em Finishing touches -- the cockpit curtain.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:cockpit_curtain}
\end{figure}

@@

The floor under this section was repaired by removing a section
of badly corroded I-beam and fabricating a new one from several
pieces of aluminum which were riveted in place to support the
floor.  Once this repair was completed, new cushion flooring was
cut and glued to the crew lounge and the washroom floor, then
various pieces of floor trim were installed.  To install the crew
seat, it was necessary to remove the lower support section, since
the fully assembled seat could not easily fit through the narrow
doorway.  Once the seat was installed, we added the crew table
and bunk bed, along with the secure storage box.  We also
installed the curtain rod and new curtains to close off the bunk
bed. The installation was finished off by re-installing the door
between the lounge and the main cabin, so that this fully
restored area can be protected from dust and debris while the
main cabin undergoes restoration over the next two years.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{new_galley_floor.eps}
   %caption of the figure 
   \caption*{\small \em New galley and floor.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:new_galley_floor}
\end{figure}
